A comparison of design-based and model-based approaches for finite population spatial data.

### 2021-12-02 inst/manuscript/manuscript.Rmd
added internal update edits

---
title: A comparison of design-based and model-based approaches for finite population spatial data.
author:
  - name: Michael Dumelle
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Jay M. Ver Hoef
    affiliation: NOAA
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Lisa Madsen
    affiliation: OSU
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
footnote:
    code: 1
    text: "Corresponding Author: Michael Dumelle (Dumelle.Michael@epa.gov)"
abstract: |
  The design-based and model-based approaches to frequentist statistical inference rest on fundamentally different foundations. In the design-based approach, inference depends on random sampling. In the model-based approach, inference depends on distributional assumptions. In this manuscript, we compare the approaches for finite population spatial data. We first provide relevant background for design-based and model-based approaches to finite population patial data. Then we use a simulation study and an analysis of real mercurcy concentration data to compare them numerically. We find that sampling plans that incorporate spatial locations (spatially balanced samples) perform better than sampling plans ignoring spatial locations (non-spatially balanced samples), regardless of whether design-based or model-based approaches were used to analyze the data. We also find that within sampling plans, model-based approaches tend to outperform design-based approaches, even for skewed data. This gap in performance is small when spatially samples are used but large when non-spatially balanced samples are used.

@@

The design-based approach can be used to select samples that are "well-spread" in space, or spatially balanced. Spatially balanced samples are useful because parameter estimates from these samples tend to vary less than parameter estimates from samples that are not spatially balanced [@stevens2004spatially; @barabesi2011sampling; @grafstrom2013well; @robertson2013bas; @wang2013design; @benedetti2017spatiallyreview]. The first spatially balanced sampling algorithm that saw widespread use was the Generalized Random Tessellation Stratified (GRTS) algorithm [@stevens2004spatially]. To quantify the spatial balance of a sample, @stevens2004spatially proposed loss metrics based on Voroni polygons. After the GRTS algorithm was developed, several other spatially balanced sampling algorithms have emerged, including the Local Pivotal Method [@grafstrom2012spatially; @grafstrom2018spatially], Spatially Correlated Poisson Sampling [@grafstrom2012spatiallypoisson], Balanced Acceptance Sampling [@robertson2013bas], Within-Sample-Distance Sampling [@benedetti2017spatially], and Halton Iterative Partitioning Sampling [@robertson2018halton]. In this manuscript, we use the Generalized Random Tessellation Stratified (GRTS) algorithm to select spatially balanced samples sampling because the algorithm has several attractive properties. It accommodates finite and infinite sampling frames. It accommodates equal, unequal, and proportional (to size) inclusion probabilities. It accommodates legacy (historical) sampling [@foster2017spatially]. It accommodates a minimum distance between units in a sample. Lastly, it accommodates replacement units in a sample, which are units that can be sampled in place of an original unit that can no longer be sampled.
The GRTS algorithm samples from finite and infinite populations by utilizing a mapping between two-dimensional and one-dimensional space. The units in the two-dimensional sampling frame are divided into cells using a hierarchical address. This hierarchical address is then used to map the units from two-dimensional space to a one-dimensional line where each unit's line length equals its inclusion probability. A systematic sample is conducted on the line and linked back to a unit in two-dimensional space, which results in the desired sample. @stevens2004spatially provides further details.

@@

After selecting a spatially balanced sample using the GRTS algorithm (i.e., a GRTS sample), data are collected and used to estimate population parameters. To unbiasedly estimate population means and totals from sample data, one can use the Horvitz-Thompson estimator [@horvitz1952generalization]. If $\tau$ is a population total, the Horvitz-Thompson estimate of $\tau$, denoted by $\hat{\tau}_{ht}$, is is given by
\begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align}
where $Z_i$ is the value of the $i$th unit in the sample and $\pi_i$ is the inclusion probability of the $i$th unit in the sample. An estimate of the population mean can be obtained by dividing $\hat{\tau}_{ht}$ by number of population units.

@@

Let $\bm{\delta} \equiv [\bm{\delta}_s \,\, \bm{\delta}_u]'$, where $\bm{\delta}_s$ and $\bm{\delta}_u$ are random errors for teh sampled and unsampled population units, respectively. We assume $E(\bm{\delta} = \mathbf{0}$ and that there is spatial correlation in $\bm{\delta}$ that can be modeled using a covariance function. It is common to assume the covariance function is second-order stationary and isotropic [@cressie1993statistics], and that the spatial covariance decreases as the separation between population units increases. Many spatial covariance functions exist, but the primary function we use throughout the simulations and applications in this manuscript is the exponential covariance function: the $i,j$th elment of hte matrix $\cov(\bm{\delta})$ is
\mbox{}
\begin{align}\label{equation:expcov}
\cov(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align}
where $\sigma^2_{1}$ is dependent random error variance measuring coarse-scale (correlated) variability, $\sigma^2_{2}$ is the independent random error variance measuring fine-scale (independent) variability, $\phi$ is the range parameter measuring the distance-decay rate of the correlation, and $h_{i,j}$ is the Euclidean distance between population units $i$ and $j$. Often $\sigma^2_{1}$ and $\sigma^2_{2}$ are called the partial sill and nugget, respectively. Any spatial covariance function could be used in the place of the exponential, including functions that allow for non-stationarity or anisotropy [@chiles1999geostatistics, pp. 80-93].

With the above model formulation, the Best Linear Unbiased Predictor (BLUP) for $f(\mathbf{b}'\mathbf{z})$ and its prediction variance can be computed. While details of the derivation are in [@verhoef2008spatial], we note here that the predictor and its variance are both moment-based, meaning that they do not rely on any distributional assumptions.

@@

Figure \ref{fig:figeff} shows the relative rMS(P)E of the four approaches from Table \ref{tab:designanalysis} using the random location layout with "IRS-Design" is the baseline. More formally, the relative rMS(P)E is defined as
\begin{equation*}
\frac{\text{rMS(P)E of sampling-analysis combination}}{\text{rMS(P)E of IRS-Design}},
\end{equation*}
When there is no spatial correlation (Figure \ref{fig:figeff}, top row), the four sampling-analysis combinations have approximately equal rMS(P)E. So, using GRTS or using a spatial model does not result in much, if any, loss in efficiency even when the response variable is not spatially correlated. When there is spatial correlation (Figure \ref{fig:figeff}, middle and bottom row), the GRTS-Model combination tends to perform best, followed by GRTS-Design, IRS-Model, and finally IRS-Design, though the difference in relative rMS(P)E among IRS-Model, GRTS-Design, and GRTS-Model is relatively small.  As the strength of spatial correlation increases, the gap in rMS(P)E between IRS-Design and the other combinations widens. Finally we note that when there is spatial correlation, IRS-Model outperforms IRS-Design by a large margin, suggesting that the poor design properties of IRS are largely mitigated by the model-based analysis. These conclusions are similar to those observed in the grid location layout. Tables for rMS(P)E in all 36 simulation scenarios are provided in the supplementary material.

@@

colour_scale <- c("goldenrod1", "goldenrod4", "mediumpurple1", "mediumpurple4")
effplot <- ggplot(data = combo_data, aes(x = n_factor, y = rel_efficiency,
                              colour = approach)) +
  facet_grid(psill_ratio ~ resptype) +
  geom_jitter(width = 0.24, size = 2.5) +
  scale_colour_manual(values = colour_scale) +
  theme_bw(base_size = 14) +
  labs(x = "Sample Size", colour = "Approach",
       y = "Relative Efficiency")
effplot
```

@@

```{r figconf, out.width = "100%", fig.cap = "Interval coverage for the four sampling-analysis combinations. The rows indicate the proportion of dependent error and the columns indicate the response type. The solid, black line in each plot represents 95\\% coverage.", message = FALSE, warning = FALSE}
ggplot(data = combo_data, aes(x = n_factor, y = coverage, colour = approach)) +
  geom_hline(yintercept = 0.95) +
  geom_jitter(width = 0.24, size = 2.5) + 
  facet_grid(psill_ratio ~ resptype) +
  scale_colour_manual(values = colour_scale) +
  ylim(0.80, 1) +
  theme_bw(base_size = 14) +
  labs(x = "Sample Size", colour = "Approach",
       y = "Interval Coverage")
```

@@

The Environmental Protection Agency (EPA), states, and tribes periodically conduct National Aquatic Research Surveys (NARS) in the United States to assess the water quality of various bodies of water. We will use the 2012 National Lakes Assessment (NLA), which measures various aspects of lake health and quality in lakes in the contiguous United States, to study mercury concentration. Although we know the true mean mercury concentration values for the 986 lakes from the 2012 NLA, we will explore whether or not we obtain an adequately precise estimate for the realized mean mercury concentration if we sample only 100 of the 986 lakes.

@@

>>>>

---
title: A comparison of design-based and model-based approaches for finite population spatial data.
author:
  - name: Michael Dumelle
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Jay M. Ver Hoef
    affiliation: NOAA
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Lisa Madsen
    affiliation: OSU
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
footnote:
    code: 1
    text: "Corresponding Author: Michael Dumelle (Dumelle.Michael@epa.gov)"
abstract: |
  The design-based and model-based approaches to frequentist statistical inference rest on fundamentally different foundations. In the design-based approach, inference depends on random sampling. In the model-based approach, inference depends on distributional assumptions. In this manuscript, we compare the approaches for finite population spatial data. We first provide relevant background for design-based and model-based approaches to finite population spatial data. Then we use a simulation study and an analysis of real mercury concentration data to compare them numerically. We find that sampling plans that incorporate spatial locations (spatially balanced samples) perform better than sampling plans ignoring spatial locations (non-spatially balanced samples), regardless of whether design-based or model-based approaches were used to analyze the data. We also find that within sampling plans, model-based approaches tend to outperform design-based approaches, even for skewed data. This gap in performance is small when spatially balanced samples are used but large when non-spatially balanced samples are used.

@@

The design-based approach can be used to select samples that are "well-spread" in space, or spatially balanced. Spatially balanced samples are useful because parameter estimates from these samples tend to vary less than parameter estimates from samples that are not spatially balanced [@stevens2004spatially; @barabesi2011sampling; @grafstrom2013well; @robertson2013bas; @wang2013design; @benedetti2017spatiallyreview]. The first spatially balanced sampling algorithm that saw widespread use was the Generalized Random Tessellation Stratified (GRTS) algorithm [@stevens2004spatially]. To quantify the spatial balance of a sample, @stevens2004spatially proposed loss metrics based on Voroni polygons. After the GRTS algorithm was developed, several other spatially balanced sampling algorithms have emerged, including the Local Pivotal Method [@grafstrom2012spatially; @grafstrom2018spatially], Spatially Correlated Poisson Sampling [@grafstrom2012spatiallypoisson], Balanced Acceptance Sampling [@robertson2013bas], Within-Sample-Distance Sampling [@benedetti2017spatially], and Halton Iterative Partitioning Sampling [@robertson2018halton]. In this manuscript, we use the Generalized Random Tessellation Stratified (GRTS) algorithm to select spatially balanced samples sampling because the algorithm has several attractive properties. It accommodates finite and infinite sampling frames, equal, unequal, and proportional (to size) inclusion probabilities, legacy (historical) sampling [@foster2017spatially], a minimum distance between units in a sample, and replacement units in a sample (which are units that can be sampled in place of an original unit that can no longer be sampled).
The GRTS algorithm samples from finite and infinite populations by utilizing a mapping between two-dimensional and one-dimensional space. The units in the two-dimensional sampling frame are divided into cells using a hierarchical address. This hierarchical address is then used to map the units from two-dimensional space to a one-dimensional line where each unit's line length equals its inclusion probability. A systematic sample is conducted on the line and linked back to a unit in two-dimensional space, which results in the desired sample. @stevens2004spatially provides further details.

@@

After selecting a spatially balanced sample using the GRTS algorithm (i.e., a GRTS sample), data are collected and used to estimate population parameters. To unbiasedly estimate population means and totals from sample data, one can use the Horvitz-Thompson estimator [@horvitz1952generalization]. If $\tau$ is a population total, the Horvitz-Thompson estimate of $\tau$, denoted by $\hat{\tau}_{ht}$, is is given by
\begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align}
where $Z_i$ is the value of the $i$th unit in the sample and $\pi_i$ is the inclusion probability of the $i$th unit in the sample. An estimate of the population mean can be obtained by dividing $\hat{\tau}_{ht}$ by the number of population units, $N$.

@@

Let $\bm{\delta} \equiv [\bm{\delta}_s \,\, \bm{\delta}_u]'$, where $\bm{\delta}_s$ and $\bm{\delta}_u$ are random errors for the sampled and unsampled population units, respectively. We assume $E(\bm{\delta}) = \mathbf{0}$ and that there is spatial correlation in $\bm{\delta}$ that can be modeled using a covariance function. It is common to assume the covariance function is second-order stationary and isotropic [@cressie1993statistics], and that the spatial covariance decreases as the separation between population units increases. Many spatial covariance functions exist, but the primary function we use throughout the simulations and applications in this manuscript is the exponential covariance function: the $i,j$th elment of the matrix $\cov(\bm{\delta})$ is
\mbox{}
\begin{align}\label{equation:expcov}
\cov(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align}
where $\sigma^2_{1}$ is dependent random error variance measuring coarse-scale (correlated) variability, $\sigma^2_{2}$ is the independent random error variance measuring fine-scale (independent) variability, $\phi$ is the range parameter measuring the distance-decay rate of the correlation, and $h_{i,j}$ is the Euclidean distance between population units $i$ and $j$. Often $\sigma^2_{1}$ and $\sigma^2_{2}$ are called the partial sill and nugget, respectively. Any spatial covariance function could be used in the place of the exponential, including functions that allow for non-stationarity or anisotropy [@chiles1999geostatistics, pp. 80-93].

With the above model formulation, the Best Linear Unbiased Predictor (BLUP) for $f(\mathbf{b}'\mathbf{z})$ and its prediction variance can be computed. While details of the derivation are in @verhoef2008spatial, we note here that the predictor and its variance are both moment-based, meaning that they do not rely on any distributional assumptions.

@@

Figure \ref{fig:figeff} shows the relative rMS(P)E of the four approaches from Table \ref{tab:designanalysis} using the random location layout with "IRS-Design" as the baseline. More formally, the relative rMS(P)E is defined as
\begin{equation*}
\frac{\text{rMS(P)E of sampling-analysis combination}}{\text{rMS(P)E of IRS-Design}},
\end{equation*}
When there is no spatial correlation (Figure \ref{fig:figeff}, top row), the four sampling-analysis combinations have approximately equal rMS(P)E. So, using GRTS or using a spatial model does not result in much, if any, loss in efficiency even when the response variable is not spatially correlated. When there is spatial correlation (Figure \ref{fig:figeff}, middle and bottom row), the GRTS-Model combination tends to perform best, followed by GRTS-Design, IRS-Model, and finally IRS-Design, though the difference in relative rMS(P)E among IRS-Model, GRTS-Design, and GRTS-Model is relatively small.  As the strength of spatial correlation increases, the gap in rMS(P)E between IRS-Design and the other combinations widens. Finally we note that when there is spatial correlation, IRS-Model outperforms IRS-Design by a large margin, suggesting that the poor design properties of IRS are largely mitigated by the model-based analysis. These conclusions are similar to those observed in the grid location layout. Tables for rMS(P)E in all 36 simulation scenarios are provided in the supplementary material.

@@

colour_scale <- c("goldenrod1", "goldenrod4", "mediumpurple1", "mediumpurple4")
resptype_labs <- c(normal = "Response: Normal", lognormal = "Response: Lognormal")
psill_ratio_labs <- c("0" = "Prop DE: 0", "0.5" = "Prop DE: 0.5", "0.9" = "Prop DE: 0.9")
effplot <- ggplot(data = combo_data, aes(x = n_factor, y = rel_efficiency,
                              colour = approach)) +
  facet_grid(
    psill_ratio ~ resptype,
    labeller = labeller(resptype = resptype_labs, psill_ratio = psill_ratio_labs)
  ) +
  geom_jitter(width = 0.24, size = 2.5) +
  scale_colour_manual(values = colour_scale) +
  theme_bw(base_size = 14) +
  labs(x = "Sample Size", colour = "Approach",
       y = "Relative Efficiency")
effplot
```

@@

```{r figconf, out.width = "100%", fig.cap = "Interval coverage for the four sampling-analysis combinations. The rows indicate the proportion of dependent error and the columns indicate the response type. The solid, black line in each plot represents 95\\% coverage.", message = FALSE, warning = FALSE}
ggplot(data = combo_data, aes(x = n_factor, y = coverage, colour = approach)) +
  geom_hline(yintercept = 0.95) +
  geom_jitter(width = 0.24, size = 2.5) + 
  facet_grid(
    psill_ratio ~ resptype,
    labeller = labeller(resptype = resptype_labs, psill_ratio = psill_ratio_labs)
  ) +
  scale_colour_manual(values = colour_scale) +
  ylim(0.80, 1) +
  theme_bw(base_size = 14) +
  labs(x = "Sample Size", colour = "Approach",
       y = "Interval Coverage")
```

@@

The Environmental Protection Agency (EPA), states, and tribes periodically conduct National Aquatic Research Surveys (NARS) in the United States to assess the water quality of various bodies of water. We will use the 2012 National Lakes Assessment (NLA), which measures various aspects of lake health and water quality in lakes in the contiguous United States, to study mercury concentration. Although we know the true mean mercury concentration values for the 986 lakes from the 2012 NLA, we will explore whether or not we obtain an adequately precise estimate for the realized mean mercury concentration if we sample only 100 of the 986 lakes.

@@

# Acknowledgements {.unnumbered}

The views expressed in this manuscript are those of the authors and do not necessarily represent the views or policies of the U.S. Environmental Protection Agency. Any mention of trade names, products, or services does not imply an endorsement by the U.S. government or the U.S. Environmental Protection Agency. The U.S. Environmental Protection Agency does not endorse any commercial products, services, or enterprises.

### 2021-12-02 inst/manuscript/manuscript.tex
added internal update edits

\usepackage{graphicx}
%%%%%%%%%%%%%%%% end my additions to header

@@

  \begin{abstract}
  The design-based and model-based approaches to frequentist statistical
  inference rest on fundamentally different foundations. In the
  design-based approach, inference depends on random sampling. In the
  model-based approach, inference depends on distributional assumptions.
  In this manuscript, we compare the approaches for finite population
  spatial data. We first provide relevant background for design-based
  and model-based approaches to finite population patial data. Then we
  use a simulation study and an analysis of real mercurcy concentration
  data to compare them numerically. We find that sampling plans that
  incorporate spatial locations (spatially balanced samples) perform
  better than sampling plans ignoring spatial locations (non-spatially
  balanced samples), regardless of whether design-based or model-based
  approaches were used to analyze the data. We also find that within
  sampling plans, model-based approaches tend to outperform design-based
  approaches, even for skewed data. This gap in performance is small
  when spatially samples are used but large when non-spatially balanced
  samples are used.
  \end{abstract}

@@

Though the design-based and model-based approaches apply to statistical
inference in a broad sense, we focus on comparing these approaches for
spatial data. We define spatial data as data that incorporates the
specific locations of the population units into either the design or
estimation process. De Gruijter and Ter Braak (1990) give an early
comparison of design-based and model-based approaches for spatial data,
quashing the belief that design-based approaches could not be used for
spatially correlated data. Since then, there have been several general
comparisons between design-based and model-based approaches for spatial
data (Brus and De Gruijter, 1997; Brus, 2021; Ver Hoef, 2002; Ver Hoef,
2008; Wang et al., 2012). Cooper (2006) reviews the two approaches in an
ecological context before introducing a ``model-assisted'' variance
estimator that combines aspects from each approach. In addition to
Cooper (2006), there has been substantial research and development into
estimators that use both design and model-based principles (see e.g.,
Sterba (2009), Cicchitelli and Montanari (2012), Chan-Golston et al.
(2020) for a Bayesian approach).

@@

After selecting a spatially balanced sample using the GRTS algorithm
(i.e., a GRTS sample), data are collected and used to estimate
population parameters. To unbiasedly estimate population means and
totals from sample data, one can use the Horvitz-Thompson estimator
(Horvitz and Thompson, 1952). If \(\tau\) is a population total, the
Horvitz-Thompson estimate of \(\tau\), denoted by \(\hat{\tau}_{ht}\),
is is given by \begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align} where \(Z_i\) is the value of the \(i\)th unit in the sample
and \(\pi_i\) is the inclusion probability of the \(i\)th unit in the
sample. An estimate of the population mean can be obtained by dividing
\(\hat{\tau}_{ht}\) by number of population units.

@@

Let \(\bm{\delta} \equiv [\bm{\delta}_s \,\, \bm{\delta}_u]'\), where
\(\bm{\delta}_s\) and \(\bm{\delta}_u\) are random errors for teh
sampled and unsampled population units, respectively. We assume
\(E(\bm{\delta} = \mathbf{0}\) and that there is spatial correlation in
\(\bm{\delta}\) that can be modeled using a covariance function. It is
common to assume the covariance function is second-order stationary and
isotropic (Cressie, 1993), and that the spatial covariance decreases as
the separation between population units increases. Many spatial
covariance functions exist, but the primary function we use throughout
the simulations and applications in this manuscript is the exponential
covariance function: the \(i,j\)th elment of hte matrix
\(\mathop{\mathrm{{cov}}}(\bm{\delta})\) is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align} where \(\sigma^2_{1}\) is dependent random error variance
measuring coarse-scale (correlated) variability, \(\sigma^2_{2}\) is the
independent random error variance measuring fine-scale (independent)
variability, \(\phi\) is the range parameter measuring the
distance-decay rate of the correlation, and \(h_{i,j}\) is the Euclidean
distance between population units \(i\) and \(j\). Often
\(\sigma^2_{1}\) and \(\sigma^2_{2}\) are called the partial sill and
nugget, respectively. Any spatial covariance function could be used in
the place of the exponential, including functions that allow for
non-stationarity or anisotropy (Chiles and Delfiner, 1999, pp. 80--93).

With the above model formulation, the Best Linear Unbiased Predictor
(BLUP) for \(f(\mathbf{b}'\mathbf{z})\) and its prediction variance can
be computed. While details of the derivation are in (Ver Hoef, 2008), we
note here that the predictor and its variance are both moment-based,
meaning that they do not rely on any distributional assumptions.

@@

Figure \ref{fig:figeff} shows the relative rMS(P)E of the four
approaches from Table \ref{tab:designanalysis} using the random location
layout with ``IRS-Design'' is the baseline. More formally, the relative
rMS(P)E is defined as \begin{equation*}
\frac{\text{rMS(P)E of sampling-analysis combination}}{\text{rMS(P)E of IRS-Design}},
\end{equation*} When there is no spatial correlation (Figure
\ref{fig:figeff}, top row), the four sampling-analysis combinations have
approximately equal rMS(P)E. So, using GRTS or using a spatial model
does not result in much, if any, loss in efficiency even when the
response variable is not spatially correlated. When there is spatial
correlation (Figure \ref{fig:figeff}, middle and bottom row), the
GRTS-Model combination tends to perform best, followed by GRTS-Design,
IRS-Model, and finally IRS-Design, though the difference in relative
rMS(P)E among IRS-Model, GRTS-Design, and GRTS-Model is relatively
small. As the strength of spatial correlation increases, the gap in
rMS(P)E between IRS-Design and the other combinations widens. Finally we
note that when there is spatial correlation, IRS-Model outperforms
IRS-Design by a large margin, suggesting that the poor design properties
of IRS are largely mitigated by the model-based analysis. These
conclusions are similar to those observed in the grid location layout.
Tables for rMS(P)E in all 36 simulation scenarios are provided in the
supplementary material.

@@

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-akaike1974new}{}%
Akaike, H., 1974. A new look at the statistical model identification.
IEEE Transactions on Automatic Control 19, 716--723.

@@

\leavevmode\hypertarget{ref-diggle2010geostatistical}{}%
Diggle, P.J., Menezes, R., Su, T., 2010. Geostatistical inference under
preferential sampling. Journal of the Royal Statistical Society: Series
C (Applied Statistics) 59, 191--232.

@@

\leavevmode\hypertarget{ref-ver2013comparison}{}%
Ver Hoef, J.M., Temesgen, H., 2013. A comparison of the spatial linear
model to nearest neighbor (k-NN) methods for forestry applications. PlOS
ONE 8, e59129.

>>>>

\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%%%%%%%%%%%%%%%% end my additions to header

@@

  \begin{abstract}
  The design-based and model-based approaches to frequentist statistical
  inference rest on fundamentally different foundations. In the
  design-based approach, inference depends on random sampling. In the
  model-based approach, inference depends on distributional assumptions.
  In this manuscript, we compare the approaches for finite population
  spatial data. We first provide relevant background for design-based and
  model-based approaches to finite population spatial data. Then we use a
  simulation study and an analysis of real mercury concentration data to
  compare them numerically. We find that sampling plans that incorporate
  spatial locations (spatially balanced samples) perform better than
  sampling plans ignoring spatial locations (non-spatially balanced
  samples), regardless of whether design-based or model-based approaches
  were used to analyze the data. We also find that within sampling plans,
  model-based approaches tend to outperform design-based approaches, even
  for skewed data. This gap in performance is small when spatially
  balanced samples are used but large when non-spatially balanced samples
  are used.
  \end{abstract}

@@

Though the design-based and model-based approaches apply to statistical
inference in a broad sense, we focus on comparing these approaches for
spatial data. We define spatial data as data that incorporates the
specific locations of the population units into either the design or
estimation process. De Gruijter and Ter Braak (1990) give an early
comparison of design-based and model-based approaches for spatial data,
quashing the belief that design-based approaches could not be used for
spatially correlated data. Since then, there have been several general
comparisons between design-based and model-based approaches for spatial
data (Brus and De Gruijter, 1997; Brus, 2021; Ver Hoef, 2002, 2008; Wang
et al., 2012). Cooper (2006) reviews the two approaches in an ecological
context before introducing a ``model-assisted'' variance estimator that
combines aspects from each approach. In addition to Cooper (2006), there
has been substantial research and development into estimators that use
both design and model-based principles (see e.g., Sterba (2009),
Cicchitelli and Montanari (2012), Chan-Golston et al. (2020) for a
Bayesian approach).

@@

After selecting a spatially balanced sample using the GRTS algorithm
(i.e., a GRTS sample), data are collected and used to estimate
population parameters. To unbiasedly estimate population means and
totals from sample data, one can use the Horvitz-Thompson estimator
(Horvitz and Thompson, 1952). If \(\tau\) is a population total, the
Horvitz-Thompson estimate of \(\tau\), denoted by \(\hat{\tau}_{ht}\),
is is given by \begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align} where \(Z_i\) is the value of the \(i\)th unit in the sample
and \(\pi_i\) is the inclusion probability of the \(i\)th unit in the
sample. An estimate of the population mean can be obtained by dividing
\(\hat{\tau}_{ht}\) by the number of population units, \(N\).

@@

Let \(\bm{\delta} \equiv [\bm{\delta}_s \,\, \bm{\delta}_u]'\), where
\(\bm{\delta}_s\) and \(\bm{\delta}_u\) are random errors for the
sampled and unsampled population units, respectively. We assume
\(E(\bm{\delta}) = \mathbf{0}\) and that there is spatial correlation in
\(\bm{\delta}\) that can be modeled using a covariance function. It is
common to assume the covariance function is second-order stationary and
isotropic (Cressie, 1993), and that the spatial covariance decreases as
the separation between population units increases. Many spatial
covariance functions exist, but the primary function we use throughout
the simulations and applications in this manuscript is the exponential
covariance function: the \(i,j\)th elment of the matrix
\(\mathop{\mathrm{{cov}}}(\bm{\delta})\) is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align} where \(\sigma^2_{1}\) is dependent random error variance
measuring coarse-scale (correlated) variability, \(\sigma^2_{2}\) is the
independent random error variance measuring fine-scale (independent)
variability, \(\phi\) is the range parameter measuring the
distance-decay rate of the correlation, and \(h_{i,j}\) is the Euclidean
distance between population units \(i\) and \(j\). Often
\(\sigma^2_{1}\) and \(\sigma^2_{2}\) are called the partial sill and
nugget, respectively. Any spatial covariance function could be used in
the place of the exponential, including functions that allow for
non-stationarity or anisotropy (Chiles and Delfiner, 1999, pp. 80--93).

With the above model formulation, the Best Linear Unbiased Predictor
(BLUP) for \(f(\mathbf{b}'\mathbf{z})\) and its prediction variance can
be computed. While details of the derivation are in Ver Hoef (2008), we
note here that the predictor and its variance are both moment-based,
meaning that they do not rely on any distributional assumptions.

@@

Figure \ref{fig:figeff} shows the relative rMS(P)E of the four
approaches from Table \ref{tab:designanalysis} using the random location
layout with ``IRS-Design'' as the baseline. More formally, the relative
rMS(P)E is defined as \begin{equation*}
\frac{\text{rMS(P)E of sampling-analysis combination}}{\text{rMS(P)E of IRS-Design}},
\end{equation*} When there is no spatial correlation (Figure
\ref{fig:figeff}, top row), the four sampling-analysis combinations have
approximately equal rMS(P)E. So, using GRTS or using a spatial model
does not result in much, if any, loss in efficiency even when the
response variable is not spatially correlated. When there is spatial
correlation (Figure \ref{fig:figeff}, middle and bottom row), the
GRTS-Model combination tends to perform best, followed by GRTS-Design,
IRS-Model, and finally IRS-Design, though the difference in relative
rMS(P)E among IRS-Model, GRTS-Design, and GRTS-Model is relatively
small. As the strength of spatial correlation increases, the gap in
rMS(P)E between IRS-Design and the other combinations widens. Finally we
note that when there is spatial correlation, IRS-Model outperforms
IRS-Design by a large margin, suggesting that the poor design properties
of IRS are largely mitigated by the model-based analysis. These
conclusions are similar to those observed in the grid location layout.
Tables for rMS(P)E in all 36 simulation scenarios are provided in the
supplementary material.

@@

\hypertarget{acknowledgements}{%
\section*{Acknowledgements}\label{acknowledgements}}
\addcontentsline{toc}{section}{Acknowledgements}

The views expressed in this manuscript are those of the authors and do
not necessarily represent the views or policies of the U.S.
Environmental Protection Agency. Any mention of trade names, products,
or services does not imply an endorsement by the U.S. government or the
U.S. Environmental Protection Agency. The U.S. Environmental Protection
Agency does not endorse any commercial products, services, or
enterprises.

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-akaike1974new}{}%
Akaike, H., 1974. A new look at the statistical model identification.
IEEE Transactions on Automatic Control 19, 716--723.

@@

\leavevmode\hypertarget{ref-diggle2010geostatistical}{}%
Diggle, P.J., Menezes, R., Su, T.-l., 2010. Geostatistical inference
under preferential sampling. Journal of the Royal Statistical Society:
Series C (Applied Statistics) 59, 191--232.

@@

\leavevmode\hypertarget{ref-ver2013comparison}{}%
Ver Hoef, J.M., Temesgen, H., 2013. A comparison of the spatial linear
model to nearest neighbor (k-nn) methods for forestry applications. PlOS
ONE 8, e59129.
